Road Segment Prioritization for Bicycle Infrastructure

### 2021-01-28 README.Rmd
Update README.Rmd
routing section edits

# Ideas and discussion

<!-- - Intro: possible trim down substantially then add some literature in (a) Calculating Potential Demand [DONE]/ (b) Routing [DONE]/ (c) Road Segment Prioritization -->
<!-- - Community Detection: Where in the document should this section be? It is now before Road Segment Prioritization, but I can merge them -->
<!-- Can go into road segment section. -->
<!-- - Should we use other cities for the sake of comparison (compare communities, road types etc) -->
- How to add tables properly (they are now added using latex code which only shows up when outputting to pdf)
- Footnotes are sometimes split between pages. How do I keep them on the same page?
<!-- - What is the best way to write pseudocode (Steps for Algorithm 1 and 2) -->

<!-- ## Missing Data -->

<!-- The following file cannot be synced to github due to its size. -->
<!-- The file is neseccary for the scripts to run. Below is a link -->
<!-- to where you can download it, and instructions on where to place it -->
<!-- in the repo file structure -->

<!-- Middle Layer Super Output Areas (December 2011) Boundaries: -->

<!--   - Source: -->
<!--     <http://geoportal.statistics.gov.uk/datasets/826dc85fb600440889480f4d9dbb1a24_0> -->
<!--   - Location in Repo: data-raw/MSOA\_2011\_Boundaries/\[Add files here\] -->

<!-- ----- -->

<!-- ## Scripts -->

<!-- The scripts should be run in the order they are numbered in (and listed -->
<!-- in here). The only exception is \_x\_dodgr\_weighting\_profiles.R. -->

<!-- Check the readme -->
<!-- [here](https://github.com/Hussein-Mahfouz/Bicycle-Network-Optimization) -->
<!-- for detailed info on each script -->

@@

*Origin-Destination (OD)* data can be obtained from a range of sources, including GPS data, household travels surveys and Census data on work locations.
In areas where observed OD data is unavailable, modelling techniques such as spatial interaction models can be used to estimate travel volumes between zones [@ref?].
In this paper we use open access data from the UK census [@ONS2011flowdata], which contains aggregate statistics on number of commuters between administrative zones --- Middle layer Super Output Areas (MSOA) --- by mode of travel. MSOAs have an average population of just over 8000 [@ofn2018population].^[
See https://wicid.ukdataservice.ac.uk/ for open access to the OD data.
]
Figures \@ref(fig:potdemhistograms) and \@ref(fig:desirefacetcycling) illustrates the proportion of trips cycled by distance and the geographic extent of the input OD dataset used in this paper.

@@

A limitation of the PCT approach outlined in the previous section is that it is reliant on external and therefore inflexible routing services. The next step in our approach is to route the potential cycling demand ($\boldsymbol{C_{p}}$) between all OD pairs onto the road network. We choose not to use the PCT approach for routing as it is reliant on external and therefore inflexible routing services.
<!-- This expands on the work of @mauttone2017bicycle, by going beyond simply favoring roads with existing cycling infrastructure to creating a hierarchy of road preference. -->

@@

The above points are accounted for by using a weighted road network for routing. This has previously been done by multiplying all road segments without cycling infrastructure by a fixed impedance factor [@mauttone2017bicycle], or by assigning different weights to the road segments proportional to the investment cost of bringing them to an acceptable level of stress for cycling  [@duthie2014optimization].

For the purposes of this paper, created a weighting profile that is adjusted to favor less stressful streets (based on information from Table  \@ref(table:osmroadtypes)), and roads with existing cycling infrastructure. We believe this to be more practical  than the simplistic approach adopted by @mauttone2017bicycle, as it goes beyond simply favoring roads with existing cycling infrastructure enabling a hierarchy of road preference based on perceived stress levels.
A wide range of weighting profiles could be used, to represent different types of cyclists and road environments, as described in @ref.
<!-- The approach is also in line with the creation of LTNs, as residential streets are those where motorized traffic is most likely to be banned in the creation of LTNs.  -->

@@

Banning cycling flow completely on trunk and primary roads may result in excessively circuitous paths, as seen in Figure \@ref(fig:boxplotcircuity). When routing using the *weighted* weighting profile in Table \@ref(table:weightprofiles), we see that shortest paths increase by less than 5\% on average from unweighted shortest paths, with the largest increases still below 30\%. When routing on primary and trunk roads is banned (*weighted_2* profile in Table \@ref(table:weightprofiles)), the average increase relative to unweighted shortest paths rises to 10\%, with certain locations experiencing more significant negative effects on accessibility.

@@

Given that cyclists will only deviate from shortest paths by a certain amount to access better cycling infrastructure (as explained in Section \@ref(introduction)), allowing flow on some stretches of trunk and primary roads is necessary to insure cycling uptake and equitable access to cycling infrastructure. In its new vision for walking and cycling, the Department for Transport acknowledges that minimal segregated stretches of bicycle lanes on main roads will be necessary to avoid circuitous cycling networks [@departmentgearchange2020].

@@

It should be reiterated that the weighting profile used for routing has been developed for the purposes of this study. It creates a hierarchy of road preference that is grounded in cyclist preferences and government plans to create LTNs. Sensitivity analysis should be done to determine an optimal weighting profile, but given the variation in city road networks<!-- (Figure \ref{fig:perc_person-km}) -->, these would probably require calibration to the specific city. More accurate routing could be carried out given the availability of road-level data. In such cases we would add additional impedance to specific roads, giving more useful routing results than the current methodology which considers all roads of the same type to be equivalent.

@@

# Community Detection

@@

A major challenge facing ‘top-down’ planning approaches is how to incorporate egalitarian principles by fairly distributing investments in cycling infrastructure. One way of quantifying this is to split up the city into smaller geospatial areas and target equal investment in each of those areas, which could also help ensure that on-the-ground surveys are made by local stakeholders, an important component of the planning process  [@parkin2018designing]. Community detection offers us a way to delineate such a split; cyclists are limited in their commuting distance (see Figure \@ref(fig:cyclinghistmanc)), and so trip attractors are more likely to have a local catchment area of cyclists.

@@

A range of algorithms could be used for prioritisation. Because policy priorities vary, we present two algorithms. Both utilize existing infrastructure from the beginning and allow us to compare a solution that focuses on utilitarianism to one that focuses on egalitarianism. In both algorithms, links are selected iteratively and the iteration at which each link is added to the solution is recorded. Investments in cycling infrastructure can be limited by budget constraints, so it can be useful to see where best to allocate a defined length of segregated infrastructure.

@@

# Overarching Policies

While segregated, connected, and direct cycling infrastructure is key to achieving high levels of cycling, research has shown that it cannot exist in a vacuum. @wardman2007factors developed a mode choice model for the UK and their results showed that improved cycling infrastructure on its own only had modest impacts on mode shift, and even the unlikely scenario of all urban routes being serviced by segregated bike lanes was forecast to increase cycling mode share by only 3%. International research shows that cities that invest in more comprehensive cycling projects have a more significant increase in the number of cyclists as well as the cycling mode share [@parkin2018designing; @pucher2010infrastructure]. These cities do not just focus on infrastructure, but on general policies as well as restricting car use. Evaluation of policies in Denmark and Germany and the Netherlands has shown that their high cycling mode share is down to a broader set of policies that also include traffic calming, filtered permeability interventions, cycling rights of way, bike parking, integration with the public transport network, and making driving cars both expensive and inconvenient [@pucher2008making]. While these policies are outside the scope of this research, it is important to recognize their key role in bringing about an increase in levels of cycling.

@@

This paper provides a methodology for prioritizing road segments for investments in cycling infrastructure, with the output being a proposed cycling network for the city under study. 
The approach aims to respect both the needs of the users and the ambitions of stakeholders working at local/regional levels.
The results, primarily detailed route network maps based on current travel behaviour derived from OD data, can provide an evidence-based starting point for introducing a direct, connected, and low-stress network for any city by leveraging available open data. Given that the “most essential activity entailed in the design of cycle-friendly infrastructure is developing a cycle network” [@parkin2018designing], we believe that the methods have great potential to inform investment in cities such as Manchester where there is political will to invest in cycling long-term. A benefit of the approach is that it has relatively modest data requirements: only the road network (from OSM), its topography (from satellite imagery), and origin-destination data (from the national census in this case) are needed, opening up the possibility of deploying the approach in other cities^[The results are easily reproducible for all UK cities, and can also be reproduced for cities elsewhere given the availability of commuter data. For instructions on reproducing the results shown in this paper, see the **README** file in the R folder of the .zip file that accompanies this paper (to be released as open source software when the paper is published).].

@@

^[The results are easily reproducible for all UK cities, and can also be reproduced for cities elsewhere given the availability of commuter data. Instructions for reproducing the results are detailed in this [Github repository](https://github.com/Hussein-Mahfouz/cycle-networks)]. -->

@@

A core part of the methodology is determining where there is a high potential for cycling, and using this as a basis for recommendations on road space reallocation in order to unlock potential cycling demand. The framework provides a routing engine to study the effects of limiting access on different road types, finding that reducing cyclist flow on roads with high through-traffic results in acceptable increases in commuting distances. As a result, a hierarchy of road preference needs to be defined to route potential cycling demand. The framework provides different algorithms to determine where investments in cycling infrastructure should be prioritized. These algorithms, based on connectivity and egalitarian principles of resource distribution, ensured that whatever the level of investment, the resulting cycling network would improve connectivity of existing cycling infrastructure.

>>>>

<!-- # Ideas and discussion -->

<!-- - Intro: possible trim down substantially then add some literature in (a) Calculating Potential Demand [DONE]/ (b) Routing [DONE]/ (c) Road Segment Prioritization -->
<!-- - Community Detection: Where in the document should this section be? It is now before Road Segment Prioritization, but I can merge them -->
<!-- Can go into road segment section. -->
<!-- - Should we use other cities for the sake of comparison (compare communities, road types etc) -->
<!-- - How to add tables properly (they are now added using latex code which only shows up when outputting to pdf) -->
<!-- - Footnotes are sometimes split between pages. How do I keep them on the same page? -->
<!-- - What is the best way to write pseudocode (Steps for Algorithm 1 and 2) -->

@@

*Origin-Destination (OD)* data can be obtained from a range of sources, including GPS data, household travels surveys and Census data on work locations.
In areas where observed OD data is unavailable, modelling techniques such as spatial interaction models can be used to estimate travel volumes between zones [@black1995spatial; @martinez2013new; @wilson1971family].
In this paper we use open access data from the UK census [@ONS2011flowdata], which contains aggregate statistics on number of commuters between administrative zones --- Middle layer Super Output Areas (MSOA) --- by mode of travel. MSOAs have an average population of just over 8000 [@ofn2018population].^[
See https://wicid.ukdataservice.ac.uk/ for open access to the OD data.
]
Figures \@ref(fig:potdemhistograms) and \@ref(fig:desirefacetcycling) illustrates the proportion of trips cycled by distance and the geographic extent of the input OD dataset used in this paper.

@@

The next step in our approach is to route the potential cycling demand ($\boldsymbol{C_{p}}$) between all OD pairs onto the road network. We choose not to use the PCT approach for routing as it is reliant on external and therefore inflexible routing services.
<!-- This expands on the work of @mauttone2017bicycle, by going beyond simply favoring roads with existing cycling infrastructure to creating a hierarchy of road preference. -->

@@

The above points are accounted for by using a weighted road network for routing. This has previously been done by multiplying all road segments without cycling infrastructure by a fixed impedance factor [@mauttone2017bicycle], or by assigning different weights to the road segments proportional to the investment cost of bringing them to an acceptable level of stress for cycling  [@duthie2014optimization].
Perceived stress has been found to be a function of vehicular traffic volume and speed [@sorton1994bicycle], both of which vary predictably with road type. 
@gehrke2020cycling build on this information to use routing impedance factors for road segments that are based on road type and the existence of cycling infrastructure.

For the purposes of this paper, we created a similar weighting profile that is adjusted to favor less stressful roads (based on the definition by @sorton1994bicycle and information from Table  \@ref(table:osmroadtypes)), and roads with existing cycling infrastructure. 
We believe this to be more practical than the simplistic approach adopted by @mauttone2017bicycle, as it goes beyond simply favoring roads with existing cycling infrastructure enabling a hierarchy of road preference based on perceived stress levels.

<!-- The approach is also in line with the creation of LTNs, as residential streets are those where motorized traffic is most likely to be banned in the creation of LTNs.  -->

@@

Banning cycling flow completely on trunk and primary roads may result in excessively circuitous paths, as seen in Figure \@ref(fig:boxplotcircuity). When routing using the *weighted* weighting profile in Table \@ref(table:weightprofiles), we see that shortest paths increase by less than 5\% on average from unweighted shortest paths, with the largest increases still below 30\%. When routing on primary and trunk roads is banned (*weighted_2* profile in Table \@ref(table:weightprofiles)), the average increase relative to unweighted shortest paths rises to 10\%, with certain locations experiencing more significant negative effects on accessibility. 
Given that cyclists will only deviate from shortest paths by a certain amount to access better cycling infrastructure (as explained in Section \@ref(introduction)), allowing flow on some stretches of trunk and primary roads is necessary to insure cycling uptake and equitable access to cycling infrastructure. In its new vision for walking and cycling, the Department for Transport acknowledges that minimal segregated stretches of bicycle lanes on main roads will be necessary to avoid circuitous cycling networks [@departmentgearchange2020].

@@

@@

It should be reiterated that the weighting profile used for routing has been developed for the purposes of this study. It creates a hierarchy of road preference that is grounded in cyclist preferences and government plans to create LTNs.
A wide range of weighting profiles could be used, to represent different types of cyclists and road environments, as described in @gehrke2020cycling and @furth2016network.
Sensitivity analysis could be done to determine an optimal weighting profile, but given the variation in city road networks<!-- (Figure \ref{fig:perc_person-km}) -->, this would probably require calibration to the specific city. More accurate routing could be carried out given the availability of road-level data. In such cases we would add additional impedance to specific roads, giving more useful routing results than the current methodology which considers all roads of the same type to be equivalent.

@@

@@

A range of algorithms could be used for prioritisation. Because policy priorities vary, we present two algorithms. Both utilize existing infrastructure from the beginning and allow us to compare a solution that focuses on utilitarianism to one that focuses on egalitarianism. In both algorithms, links are selected iteratively and the iteration at which each link is added to the solution is recorded. Investments in cycling infrastructure can be limited by budget constraints, so it can be useful to see where best to allocate a defined length of segregated infrastructure. In order to incorporate egalitarian principles in our approach, we use community detection to partition the study area and evaluate the distribution of investment over the different subdivisions.

## Community Detection

A major challenge facing ‘top-down’ planning approaches is how to incorporate egalitarian principles by fairly distributing investments in cycling infrastructure. 
One way of quantifying this is to split up the city into smaller geospatial areas and target equal investment in each of those areas. This approach could also help ensure that on-the-ground surveys are made by local stakeholders, an important component of the planning process  [@parkin2018designing]. 
Community detection offers us a way to delineate such a split; cyclists are limited in their commuting distance (see Figure \@ref(fig:cyclinghistmanc)), and so trip attractors are more likely to have a local catchment area of cyclists.

@@

@@

@@

This paper provides a methodology for prioritizing road segments for investments in cycling infrastructure, with the output being a proposed cycling network for the city under study. 
The approach aims to respect both the needs of the users and the ambitions of stakeholders working at local or regional levels.
The results, primarily detailed route network maps based on current travel behaviour derived from OD data, can provide an evidence-based starting point for introducing a direct, connected, and low-stress network for any city by leveraging available open data. Given that the “most essential activity entailed in the design of cycle-friendly infrastructure is developing a cycle network” [@parkin2018designing], we believe that the methods have great potential to inform investment in cities such as Manchester where there is political will to invest in cycling long-term. A benefit of the approach is that it has relatively modest data requirements: only the road network (from OSM), its topography (from satellite imagery), and origin-destination data (from the national census in this case) are needed, opening up the possibility of deploying the approach in other cities^[The results are easily reproducible for all UK cities, and can also be reproduced for cities elsewhere given the availability of commuter data. For instructions on reproducing the results shown in this paper, see the **README** file in the R folder of the .zip file that accompanies this paper (to be released as open source software when the paper is published).].

@@

^[The results are easily reproducible for all UK cities, and can also be reproduced for cities elsewhere given the availability of commuter data. Instructions for reproducing the results are detailed in this [Github repository](https://github.com/Hussein-Mahfouz/cycle-networks)]. -->
A core part of the methodology is determining where there is a high potential for cycling, and using this as a basis for recommendations on road space reallocation in order to unlock potential cycling demand. The framework provides a routing engine to study the effects of limiting access on different road types, finding that reducing cyclist flow on roads with high through-traffic results in acceptable increases in commuting distances. As a result, a hierarchy of road preference needs to be defined to route potential cycling demand. The framework provides different algorithms to determine where investments in cycling infrastructure should be prioritized. These algorithms, based on connectivity and egalitarian principles of resource distribution, ensured that whatever the level of investment, the resulting cycling network would improve connectivity of existing cycling infrastructure.

@@

The approach is also focused solely on the allocation of cycling infrastructure, and does not consider the larger political and regulatory environment that needs to exist to promote cycling; while segregated, connected, and direct cycling infrastructure is key to achieving high levels of cycling, research has shown that it cannot exist in a vacuum. 
@wardman2007factors developed a mode choice model for the UK and their results showed that improved cycling infrastructure on its own only had modest impacts on mode shift, and even the unlikely scenario of all urban routes being serviced by segregated bike lanes was forecast to increase cycling mode share by only 3%. 
International research shows that cities that invest in more comprehensive cycling projects have a more significant increase in the number of cyclists as well as the cycling mode share [@parkin2018designing; @pucher2010infrastructure]. These cities do not just focus on infrastructure, but on general policies as well as restricting car use. Evaluation of policies in Denmark and Germany and the Netherlands has shown that their high cycling mode share is down to a broader set of policies that also include traffic calming, filtered permeability interventions, cycling rights of way, bike parking, integration with the public transport network, and making driving cars both expensive and inconvenient [@pucher2008making]. 
While these policies are outside the scope of this research, it is important to recognize their key role in bringing about an increase in levels of cycling.
